---
title: "Module 1: Climate change"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

```{r}
#| warning: false
library(tidyverse)
library(ggpmisc)
library(readr)
library(zoo)
library(ggbreak)
```

### Use of GitHub

Link to your forked GH repository:

### Use of Quarto

Link to your .qmd file:

## CO2 trends from Mauna Loa

We will begin at the global scale by plotting the famed Mauna Loa CO~2~ curve at monthly intervals, with the twin objectives of (1) making layered plots with 'ggplot2', a package within the 'tidyverse' collection widely adopted by the R data science community, and (2) refreshing understanding of the seasonal cycle in atmospheric \[CO~2~\].

With data you've collected yourself, it is more common to store locally as a csv. However, we can use tidyverse tools to read directly from files stored online and will do so here to get the latest data.

```{r}
#| echo: true
#| warning: false
#| output: false
co2 <-read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date", "average",
                                "interpolated", "trend", "days"),
                  na = c("-1", "-99.99"))
```

You can examine these data in several ways. One option is to use your mouse in the Environment tab and click on the dataset name. Alternatively, you can type `View(co2)` in your console to achieve the same outcome. Below are some other common options to understand the data structure at a glance.



### Q1 (1 point)

Create a timeseries plot of average \[CO~2~\]. Use `as.Date()` to create a formatted date for the x-axis. Label both the y-axis and add a line to connect the points.


```{r}
#| echo: true

co2_filtered = co2


co2_filtered |> ggplot(aes(x = date_decimal(decimal_date), 
                           y = average )) +
  geom_point(size = .5) + 
  geom_line( linewidth = .05)+
  theme_classic() + 
  xlab("Date of Measurement") + 
  ylab("Average CO2 level ( PPM )")

```

I don't think geom_line is very helpful here; in fact I would rather not use it at all. 



### Q2 (1 point)

In which 3-month window are the \[CO~2~\] values at their seasonal peak? Create another dataframe with these months selected and create a plot like above, but with the maximum months colored in "forestgreen".

```{r}

co2_max_months = co2 |> 
  group_by(year) |> 
  arrange(desc(average), .by_group = TRUE)|>
  mutate(max_month = max(average),
         second_max = average[2],
         third_max = average[3]) |> 
  select(max_month,
         second_max,
         third_max,
         year,month,
         average, decimal_date) 
  


#co2_max_months |> 
 # filter(average == max_month | 
  #         average == second_max | 
   #        average == third_max )

# it seems like april, may, and june 
#have the highest CO2 measurements across the dataset.

 


co2_max_months |> 
  ggplot(aes(x = date_decimal(decimal_date), 
             y = average, 
             color = as.factor(month), group = 1)) + 
  geom_point(size = .5) + 
  scale_color_manual(values = c("4" = "forestgreen", 
                                "5" = "forestgreen", 
                                "6" = "forestgreen"))+
  theme_classic() + 
  geom_line(linewidth = .05)+
  xlab("Date of Measurement") +
  ylab("Average CO2 levels ( PPM )") +
  labs(color = "Months with Most Seasonal CO2")

#point and line sizes are low so that they look better in the rendered pdf 
```



### Q3 (1 point)

Why does \[CO~2~\] peak in the months highlighted above? Consider the location of Mauna Loa and the seasonality of processes that emit and absorb \[CO~2~\].  

After reading Lewandowsky et al. 2016, how might you go about separating the long term trend and the seasonal variation in \[CO~2~\]?


  Mauna Loa is in Hawaii, and April, May, and June mark the start of the Hawaiian summer. Under warmer conditions, it is less likely for CO2 to be disolved in the ocean or any body of water, meaning its partial pressure in the surrounding atmosphere will increase slightly. At the same time, the local plants will have just emerged from the "winter" months, meaning they are probably less abundant than in late summer. Higher volumes of plant mass will consume more CO2, which would explain why average CO2 levels begin dropping right after they peak in april.

  Lewandowsky and associates separated the overall CO2 data into smaller 15-year trends and then calculated the average rate of CO2 increase in each window. Then they plotted each average against the total average of every CO2 measurement from 1970 onwards. We could apply a similar strategy here; since we care about seasonal variation and total variation, we could plot each month's average against the yearly average, then do this for each year in the data set, thereby making the overall trend clear while still illustrating that in a given year, the measurements might dance around the overall trend. 



## Global temperature

Next, we turn to a global temperature anomaly [dataset](https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt) from NASA GISS, similar to the GMST used by Lewandowsky et al. 2016.

### Q4 (1 point)

Read some [background](https://data.giss.nasa.gov/gistemp/history/background.html) as well as the [simplified documentation](https://climate.nasa.gov/vital-signs/global-temperature/?intent=121) for this dataset.

How are the measurements made? Describe each column in the dataset and its units. What is the temporal resolution of the data?

```{r}

Nasa_GISS_stuff = read_table("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt",
                             col_names = c("year",
                                           "avg_temp_anomaly_by_year",
                                           "smoothed_anomaly"))


Nasa_GISS_stuff = Nasa_GISS_stuff


```


There are three columns in the GISS global temp dataset. The first records the year the measurements were taken, the second describes the absolute mean of all temperature anomaly measurements taken in a given year and is expressed as a difference in comparison to the average surface temperature of the 1951 - 1980 time period, and the third represents the smoothed version of the second column's data using the Lowess regression method. Both the second and third column are in units of degrees Celsius.


Surface temperature measurements were taken as follows: monitoring stations were grouped in 80 surface area boxes, each with equal area, and each station would obtain air temperature anomaly series for its box. Boxes that contained areas of ocean estimated air temperature from the water temperature recorded by boats / buoys. Then the station would condense all series into a single one, which was averaged with the other stations that occupied the same one of eight different latitude sections. The global mean was then obtained by averaging the eight latitude sections. 



### Q5 (1 point)

Construct code to import the dataset from the above url in a format that can be used for transforming and visualization.

```{r}
Nasa_GISS_stuff = read_table("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt",
                             col_names = c("year",
                                           "avg_temp_anomaly_by_year",
                                           "smoothed_anomaly"))


Nasa_GISS_stuff = Nasa_GISS_stuff[-c(1:4), ] 

Nasa_GISS_stuff|> summarise(min = min(avg_temp_anomaly_by_year), max = max(avg_temp_anomaly_by_year))
```


### Q6 (1 point)

Plot the trend in global mean temperature over time. Describe what you see and how you interpret the patterns you observe.

```{r}

Nasa_GISS_stuff |>
  arrange(desc(as.numeric(avg_temp_anomaly_by_year))) |>
  ggplot(aes(x = year,
             y = as.numeric(avg_temp_anomaly_by_year), group = 1)) +
  theme_classic()+
  geom_line(data = Nasa_GISS_stuff, 
            aes(x = year, 
                y = as.numeric(smoothed_anomaly),
                linewidth = .5), 
            color = "cyan")+
  geom_point()+
  geom_line()+
  labs(x = "Year of Measurement", 
       y = "Global Temp Anomaly Compared to 1951 - 1981 Average Temp ( C )", 
       linewidth = "Overall Smoothed Trend")+
  scale_x_discrete(breaks = c(1880,
                                1900,
                                1920,
                                1940,
                                1960,
                                1980,
                                2000,
                                2020))
  
  
```

The global yearly average temperature anomalies fluctuate about the smoothed trend. Low / high temperatures seem to cycle back and forth to low / high temps every 5 - 7 years. Beginning around 1970, this cycle continues however every local low / high temperature maximum is greater than in the previous cycle. The smooth trend reflects these cycles in that there are regions where fast change in temp anomaly occurs, followed by slow changes in temp anomaly.

### Q7 (1 point)

Consider the "pause" analyzed in Lewandowsky et al. 2016. Do you see evidence for such a pause?

For there to be a pause, the rate of change of temperature anomaly would have to be zero, which would be indicated by any horizontal regions of a curve. If we examine the smoothed trend from 1970 onwards, there exist no such horizontal areas, and furthermore, the slopes of every region of the curve remain positive, meaning warming has continued, and the slopes of the later areas of the curve ( ~1990 and onwards ) are steeper and steeper, meaning warming is accelerating, not slowing down. 



### Bonus 1 (1 point)

Recreate the analysis from Lewandowsky's Fig. 1.

## Long term records of CO2

The data from Mauna Loa go back as far as modern instrumentation. To put these values into geological perspective, we must go back much further than humans have been monitoring \[CO~2~\]. Ice sheets contain layers of snow with trapped air bubbles and can be cored and dated like tree rings.

Previously hosted by Oak Ridge National Lab, the [data](https://zenodo.org/records/3678928) associated with the famed Vostok Ice Core is now archived on Zenodo. Skim [Barnola et al. 1987](https://www.nature.com/articles/329408a0) as needed.

### Q8 (1 point)

After downloading this dataset, create a folder entitled `data` and save the file within the folder with a .txt extension. Read this file into R. What are the columns and units? In the "Vostok CO2 record" section of Barnola et al. 1987, what is the temporal resolution of the data?


```{r}


ice_core = read_table("C:/Users/dante/OneDrive/Documents/Module1-climate/vostok.icecore.txt", 
                      col_names = c("depth(m)","ice_age(yr)","air_age(yr)","co2_conc(ppmv)"))

ice_core = ice_core[-c(1:20), ]

```

Column 1 represents the depth in meters in the ice core that a particular measurement was taken. Column two indicates the age in years of the ice at the depth the measurement was taken, while column three indicates the age of the air recovered from the ice at that depth in years. Column 4 reports the concentration of CO2 present in the air sample recovered in parts per million. In Barnola et. al, the time scale dates from present day to around 160k years ago with each measurement corresponding to about 2000 years of history, however in this dataset, the time scale extends to around 420k years ago.


### Q9 (1 point)

Transform the data as needed to create a plot in chronological order. Add at least two different smoothing window averages to your plot. How did you select the smoothing windows?


```{r}

options(scipen = 999)


#get three point average smoothing
ice_core = ice_core |> 
  mutate(three_point_avg = rollmean(x = as.numeric( `co2_conc(ppmv)`), 
                                                         k = 3,
                                                         fill = NA, 
                                                         align = "center"))

# get lowess smoothing 
                      
LOWESS_core = lowess(x = as.numeric(ice_core$`air_age(yr)`),
                     y = as.numeric(ice_core$`co2_conc(ppmv)`), 
                     f = 10, 
                     iter = 2,
                     delta = .01*diff(range(as.numeric(ice_core$`air_age(yr)`))))         
           
#no way i'm binding the LOWESS df to the ice_core df         

#renaming it so column names don't confuse ggplot
LOWESS_core = as_tibble(LOWESS_core) |>
  rename(air_age_smooth = x, 
         co2_conc_smooth = y)

# plot of depth vs ice age 
ice_core |> ggplot(aes(x = as.numeric( `ice_age(yr)`), 
                       y = as.numeric( `depth(m)`))) + 
  geom_point() +
  scale_y_continuous(transform = "reverse")+
  theme_classic()+
  labs(x = "Age of Ice ( yr )", y = "Depth of Measurement ( m )")

# plot of co2 conc. over time with smoothed lines

ice_core |> ggplot(aes(y = as.numeric(`co2_conc(ppmv)`), 
                       x = as.numeric( `air_age(yr)`))) + 
  geom_point(color = "white") +
  geom_line(color = "white")+
  geom_line(data = LOWESS_core, 
            aes(x = air_age_smooth, 
                y = co2_conc_smooth, 
                color = "LOWESS Regression"))+
  geom_line(data = ice_core, 
            aes(x = as.numeric(`air_age(yr)`), 
                y = as.numeric(three_point_avg), 
                color = "Three-Point Average"), linewidth = 1.5)+
  scale_x_continuous(transform = "reverse")+
  theme_classic()+
  theme(panel.background = element_rect(fill = "black"), 
        plot.background = element_rect(fill = "black"),
        axis.title = element_text(color = "white"),
        axis.text = element_text(color = "white"),
        axis.ticks = element_line(color = "white"))+
  labs(x = "Age of Air ( years before present )", 
       y = "CO2 Concentration ( ppm )", 
       color = "Smoothing Average Method")


```

I would like to sepcify that I had no idea about smoothing averages before starting this homework, so the ones I'm using here are the LOWESS regression because it was featured in the NASA GISS global temp dataset and the simple moving average because it was the first thing that popped up when I googled what a smoothing average was. 

For the smoothing windows, I changed the number of points for the point average and looked a what came out, and it seems like altering the number of points included in each sweep doesn't really change the shape of the curve that much, so I went with a 3 point average because the curve began as close to the start of the data set as possible. The same is true for the LOWESS regression; changing the smooth span value does not have much of an effect on the shape of the curve, so I went with 10 because that was my initial entry.

### Q10 (1 point)

Join your data with the Mauna Loa data and plot together, considering different colors for the different datasets.



```{r}


co2_filtered_average = co2_filtered |>
  group_by(year) |> 
  mutate(yearly_avg = mean(average))|>
  select(yearly_avg,year) |>
  unique()



ice_core |> ggplot(aes(y = as.numeric(`co2_conc(ppmv)`),
                       x = -as.numeric(`air_age(yr)`), 
                       color = "Vostok Ice Core Data"))+ 
  geom_point()+ 
  geom_line()+
  coord_cartesian(xlim = c(-430000, 1960))+
  scale_x_break(c(0,1960), scales = .5)+
  geom_point(data = co2_filtered_average,
         aes(x = year,
             y = yearly_avg, color = "Mauna Loa Station Data"))+
  geom_line(data = co2_filtered_average,
         aes(x = year,
             y = yearly_avg, color = "Mauna Loa Station Data"))+
  labs(x = "Year", y = "Co2 Concentration ( ppm )", color = "Dataset Used")+
  theme_classic()

  

```




### Bonus 2 (1 point)

Write a catchy title and short description in the style of a NYtimes infographic, contrasting the rate of anthropogenic vs. background climate change for a general public audience.




